---
title: "Smoothing - Health"
author: "Team Health"
date: "12/2/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(sp) 
library(gstat) 
library(tmap)
library(sf)
library(tidyverse)
library(magrittr)
library(rgdal)
library(flextable)
library(gridExtra)

knitr::opts_chunk$set(echo = F, message = F, warning = F)
options(scipen=200, digits=2)
```

# Import Data - Tract Level
```{r }
health <- st_read("smoothing_files/PLACES_ Census Tract Data (GIS Friendly Format), 2020 release/geo_export_735ca527-23eb-442e-bbd4-4dacf4ff0a25.shp",quiet = TRUE)
#censustract <- st_read("/Users/karidachai/Desktop/Fall 2021/MA675/tree/Census2020_Tracts/Census2020_Tracts.shp",quiet = TRUE)
asthma <- health %>% dplyr::select(stateabbr, statedesc, countyname, countyfips, tractfips, geometry, totalpopul, casthma_cr)
mental <- health %>% dplyr::select(stateabbr, statedesc, countyname, countyfips, tractfips, geometry, totalpopul, mhlth_crud)

asthma_ = subset(asthma, countyname == "Suffolk"& stateabbr == "MA")
mental_ = subset(mental, countyname == "Suffolk"& stateabbr == "MA")

tract<- st_read("smoothing_files/Census2020_Tracts/Census2020_Tracts.shp",quiet = TRUE)

asthma_$TRACTCE10 <- substring(asthma_$tractfips,6, nchar(asthma_$tractfips))
asthma2 <- asthma_[asthma_$TRACTCE10 %in% tract$TRACTCE20,]

mental_$TRACTCE10 <- substring(mental_$tractfips,6, nchar(mental_$tractfips))
mental2 <- mental_[mental_$TRACTCE10 %in% tract$TRACTCE20,]

tract$Asthma[tract$TRACTCE20 %in% asthma2$TRACTCE10] <- asthma2$casthma_cr [asthma2$TRACTCE10 %in% tract$TRACTCE20]
#tract$Asthma[is.na(tract$Asthma)] <- 0
tract$Asthma <- log(tract$Asthma)
tm_shape(tract) +
    tm_polygons(col ="Asthma",palette = "Reds", style = "fixed", breaks = seq(2.2,2.6, 0.1),
                legend.hist = TRUE) +
    tm_layout(legend.outside = TRUE)



tract$Mental[tract$TRACTCE20 %in% mental2$TRACTCE10] <- mental2$mhlth_crud [mental2$TRACTCE10 %in% tract$TRACTCE20]
#tract$Mental[is.na(tract$Mental)] <- 0
tract$Mental <- log(tract$Mental)
tm_shape(tract) +
    tm_polygons(col ="Mental",palette = "Blues",style = "fixed", breaks = seq(2.2,2.6, 0.1),
                legend.hist = TRUE) +
    tm_layout(legend.outside = TRUE)

```

# Function
```{r}
#Spherical and Gaussian Variogram functions
spherical_variogram <- function (n, ps, r) function (h) {
  h <- h / r
  n + ps * ifelse(h < 1, 1.5 * h - .5 * h ^ 3, 1)
}
gaussian_variogram <- function (n, ps, r)
  function (h) n + ps * (1 - exp(-(h / r) ^ 2))
exponential_variogram <- function (n, ps, r)
  function (h) n + ps * (1 - exp(-(h / r)))
# solves `A * x = v` where `C = chol(A)` is the Cholesky factor:
chol_solve <- function (C, v) backsolve(C, backsolve(C, v, transpose = TRUE)) # the solution of triangular system.The result will be a vector if 'x' is a vector and a matrix if 'x' is a matrix.

#New Kriging Function
epsilon <- 1e-3
kriging_smooth_spherical <- function (formula, data, ...) {
  v <- variogram(formula, data)
  v_fit <- fit.variogram(v, vgm("Sph",...))
  v_fit$psill[1] <- max(v_fit$psill[1], epsilon)
  v_f <- spherical_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2])
  
  Sigma <- v_f(as.matrix(dist(st_coordinates(data)))) # semivariogram
  Sigma <- sum(v_fit$psill) - Sigma # prior variance
  tau2 <- v_fit$psill[1] # residual variance
  C <- chol(tau2 * diag(nrow(data)) + Sigma)
  y <- model.frame(formula, data)[, 1] # response
  x <- model.matrix(formula, data)
  # generalized least squares:
  xt <- backsolve(C, x, transpose = TRUE)
  beta <- coef(lm.fit(xt, backsolve(C, y, transpose = TRUE))) # prior mean
  names(beta) <- colnames(x)
  beta_se <- sqrt(diag(chol2inv(chol(crossprod(xt)))))
  
  chol_sigma <- chol(Sigma)
  Sigma_inv <- chol2inv(chol_sigma)
  C <- chol(Sigma_inv + diag(nrow(data)) / tau2)
  # posterior mean (smoother):
  mu <- drop(chol_solve(C, y / tau2 + Sigma_inv %*% x %*% beta))
  
  sinv_mu <- chol_solve(chol_sigma, mu - x %*% beta)
  krige <- function (new_data) { # prediction function
    D <- apply(coordinates(data), 1,
               function (coord) apply(st_coordinates(new_data), 1, dist2, coord))
    V <- sum(v_fit$psill) - v_f(D)
    
    t <- delete.response(terms(formula))
    xp <- model.matrix(t, model.frame(t, new_data))
    drop(xp %*% beta + V %*% sinv_mu)
  }
  
  list(smooth = mu, prior_coef = beta, prior_coef_se = beta_se,
       variogram = v_fit, krige = krige)
}
```

# Original Asthma

```{r}
tract2<-st_centroid(tract) #Center the polygon
tract2 = na.omit(tract2)
#tract2$a = exp(tract2$Asthma)
tract2$a = (tract2$Asthma)
tm_shape(tract2) +
  tm_bubbles(col = "Asthma", palette = "-RdYlBu", size = .3, alpha = .5)
v <- variogram(a ~ 1, tract2)
plot(v, plot.numbers = T)
v_fit <- fit.variogram(v, vgm("Sph")) #'vgm':generate a variogram model or add to an existing model.
v_f <- spherical_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2])

# check variogram and covariance
op <- par(mfrow = c(1, 2))
h <- seq(0, 40000, length = 100)
plot(v$dist, v$gamma,  pch = 19, col = "gray",
     xlab = "distance", ylab = "semivariogram")
lines(h, v_f(h))
abline(v = v_fit$range[2], col = "gray")
plot(h, sum(v_fit$psill) - v_f(h), type = "l",
     xlab = "distance", ylab = "covariogram",
     ylim = c(0, sum(v_fit$psill)))
points(0, sum(v_fit$psill), pch = 19)
abline(v = v_fit$range[2], col = "gray")
par(op)

#coordinates- x & y
#tract3 <- tract2 %>%
#    mutate(x = unlist(map(tract2$geometry,1)),
#           y = unlist(map(tract2$geometry,2)))
#tract3

#tract4 <- tract3 %>% st_sf() %>% as_Spatial()

# check smoothing
ks <- kriging_smooth_spherical(a ~ 1, tract2)
y <- tract2$a
op <- par(mfrow = c(1, 2))
plot(ks$smooth, y, ylim = c(2, 3)); abline(0, 1, col = "red")
plot(ks$smooth, type = "l", ylab = "y")
points(y, pch = 19, col = "gray")
abline(h = ks$prior_mean)
par(op)
tract2$smootha <- ks$smooth
breaks <- seq(4.5, 8, by = .5)
tmap_mode("plot")

tmap_arrange(
  tm_shape(tract2) +
    tm_bubbles(col = "Asthma", palette = "-RdYlBu", size = .3, breaks = breaks)
  #tm_shape(tract2) +
    #tm_bubbles(col = "smootha", palette = "-RdYlBu", size = .3, breaks = breaks)
)

```

# Original Mental Health

```{r}
tract2$b = (tract2$Mental)
tm_shape(tract2) +
  tm_bubbles(col = "Mental", palette = "-RdYlBu", size = .3, alpha = .5)
v <- variogram(b ~ 1, tract2)
plot(v, plot.numbers = T)
v_fit <- fit.variogram(v, vgm("Sph")) #'vgm':generate a variogram model or add to an existing model.
v_f <- spherical_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2])

# check variogram and covariance
op <- par(mfrow = c(1, 2))
h <- seq(0, 40000, length = 100)
plot(v$dist, v$gamma,  pch = 19, col = "gray",
     xlab = "distance", ylab = "semivariogram")
lines(h, v_f(h))
abline(v = v_fit$range[2], col = "gray")
plot(h, sum(v_fit$psill) - v_f(h), type = "l",
     xlab = "distance", ylab = "covariogram",
     ylim = c(0, sum(v_fit$psill)))
points(0, sum(v_fit$psill), pch = 19)
abline(v = v_fit$range[2], col = "gray")
par(op)

#coordinates- x & y
#tract3 <- tract2 %>%
#    mutate(x = unlist(map(tract2$geometry,1)),
#           y = unlist(map(tract2$geometry,2)))
#tract3

#tract4 <- tract3 %>% st_sf() %>% as_Spatial()

# check smoothing
ks <- kriging_smooth_spherical(b ~ 1, tract2)
y <- tract2$b
op <- par(mfrow = c(1, 2))
plot(ks$smooth, y); abline(0, 1, col = "red")
plot(ks$smooth, type = "l", ylab = "y")
points(y, pch = 19, col = "gray")
abline(h = ks$prior_mean)
par(op)
tract2$smoothb <- ks$smooth
breaks <- seq(4.5, 8, by = .5)
tmap_mode("plot")

tmap_arrange(
  tm_shape(tract2) +
    tm_bubbles(col = "Mental", palette = "-RdYlBu", size = .3, breaks = breaks)
  #tm_shape(tract2) +
    #tm_bubbles(col = "smooth", palette = "-RdYlBu", size = .3, breaks = breaks)
)
```

# Merge data with enviornment group

```{r}
environ <- st_read("smoothing_files/Tree_dem__env_1129/Tree_dem_env.shp",quiet = TRUE)

library(tidyverse)
newhealth <- environ %>% inner_join(as.data.frame(tract2)[,-16], by  = 'GEOID20')
newhealth <- newhealth%>%st_as_sf(sf_column_name = "geometry")
```

# Model - Asthma

```{r}
smoothing_formula <- smootha ~ MnrPrc_s + MnrPrc_r + MdGrTx_s + MdGrTx_r + Mdn_Inc_s + Mdn_Inc_r + mn_pm25_s + mn_pm25_r + mn_tmp_s + mn_tmp_r + Tre_Cnp
v <- variogram(smoothing_formula, newhealth)

```

```{r}
v_fit <- fit.variogram(v, vgm("Wav"))
v_f <- spherical_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2])
h <- seq(0, 200, length = 1000) 
plot(v$dist, v$gamma,  pch = 19, col = "gray",
     xlab = "distance", ylab = "semivariogram", main = "Asthma")
lines(h, v_f(h))
abline(v = v_fit$range[2], col = "gray")


# check smoothing:
ks_as <- kriging_smooth_spherical(smoothing_formula, newhealth)
y <- newhealth$smootha
op <- par(mfrow = c(1, 2))
plot(ks_as$smooth, y, ylim = c(2.3, 2.5)); abline(0, 1, col = "red") 
plot(ks_as$smooth, type = "l", ylab = "y")
points(y, pch = 19, col = "gray") 
abline(h = ks_as$prior_coef) 
par(op)
```

### Model coefficient - Asthma

```{r}
library(flextable)
coef1 <- as.data.frame(t(rbind(ks_as$prior_coef, ks_as$prior_coef_se)))
coef1 <- cbind(row.names(coef1),coef1) 
names(coef1) <- c("Predictors", "Coefficient", "Sd.")
flextable(coef1) %>% theme_booktabs() %>% autofit()
```

### Significant Model coefficient - Asthma

```{r}
coef1$Q2.5 <- coef1$Coefficient - 2 * coef1$Sd.
coef1$Q97.5 <- coef1$Coefficient + 2 * coef1$Sd.
coef1 <- coef1[(coef1$Q97.5 > 0 & coef1$Q2.5 >0)| (coef1$Q97.5 < 0 & coef1$Q2.5 < 0),]
names(coef1) <- c("Predictor", "Coeffient", "Sd.", "Q2.5", "Q97.5")
flextable(coef1) %>% theme_booktabs() %>% autofit()
```

## Refit model for significant predictors - Asthma

```{r}
smoothing_formula2 <- smootha ~ MdGrTx_s + MdGrTx_r + Mdn_Inc_s + Mdn_Inc_r + Tre_Cnp
v <- variogram(smoothing_formula2, newhealth)
```

```{r}
v_fit <- fit.variogram(v, vgm("Wav"))
v_f <- spherical_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2])
h <- seq(0, 200, length = 1000) 
plot(v$dist, v$gamma,  pch = 19, col = "gray",
     xlab = "distance", ylab = "semivariogram", main = "Asthma_Refit")
lines(h, v_f(h))
abline(v = v_fit$range[2], col = "gray")


# check smoothing:
ks_as_refit <- kriging_smooth_spherical(smoothing_formula2, newhealth)
y <- newhealth$smootha
op <- par(mfrow = c(1, 2))
plot(ks_as_refit$smooth, y, ylim = c(2.3, 2.5)); abline(0, 1, col = "red") 
plot(ks_as_refit$smooth, type = "l", ylab = "y")
points(y, pch = 19, col = "gray") 
abline(h = ks_as_refit$prior_coef) 
par(op)
```

### Model coefficient - Asthma

```{r}
coef1 <- as.data.frame(t(rbind(ks_as_refit$prior_coef, ks_as_refit$prior_coef_se)))
coef1 <- cbind(row.names(coef1),coef1) 
```

```{r}
coef1$Q97.5 <- coef1$V1 + 2 * coef1$V2
coef1$Q2.5 <- coef1$V1 - 2 * coef1$V2
names(coef1) <- c("Predictor", "Coeffient", "Sd.", "Q2.5", "Q97.5")
flextable(coef1) %>% theme_booktabs() %>% autofit()
```

# After-Smoothing Map & Kriging Predicted Map for Asthma

```{r}
tract7 <- st_read("smoothing_files/matches/matches.shp", quiet = T)
newhealth$ks_as_refit <- ks_as_refit$smooth
tract7$ks_as_refit[tract7$GEOID20 %in% newhealth$GEOID20] <- newhealth$ks_as_refit[newhealth$GEOID20 %in% tract7$GEOID20]

ggplot() + geom_sf(color = 'grey', data = tract7) + geom_sf(aes(fill = ks_as_refit), data=na.omit(tract7)) + 
  scale_fill_distiller(palette="Reds", trans = "reverse") + labs(title = 'Original Model for Asthma')
grid <- st_centroid(tract7[is.na(tract7$ks_as_refit),]) 
pred_AS <-  krige(ks_as_refit ~1, newhealth, grid$geometry, model = v_fit)
tract7$ks_as_refit[is.na(tract7$ks_as_refit)] <- pred_AS$var1.pred

ggplot(data = tract7) + geom_sf(color = 'grey') + geom_sf(aes(fill = ks_as_refit)) + 
  scale_fill_distiller(palette="Reds", trans = "reverse") + labs(title = 'Refitted Model for Asthma')

```


# Model Mental Health
```{r}
smoothing_formula1 <- smoothb ~ MnrPrc_s + MnrPrc_r + MdGrTx_s + MdGrTx_r + Mdn_Inc_s + Mdn_Inc_r + mn_pm25_s + mn_pm25_r + mn_tmp_s + mn_tmp_r + Tre_Cnp
v2 <- variogram(smoothing_formula1, newhealth)
```

```{r}
v_fit <- fit.variogram(v2, vgm("Sph"))
v_f <- spherical_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2])
h <- seq(0, 200, length = 1000) 
plot(v2$dist, v2$gamma,  pch = 19, col = "gray",
     xlab = "distance", ylab = "semivariogram", main = "Mental Health")
lines(h, v_f(h))
abline(v2 = v_fit$range[2], col = "gray")


# check smoothing:
ks_mh <- kriging_smooth_spherical(smoothing_formula1, newhealth)
y <- newhealth$smoothb
op <- par(mfrow = c(1, 2))
plot(ks_mh$smooth, y, ylim = c(2.2,3)); abline(0, 1, col = "red") 
plot(ks_mh$smooth, type = "l", ylab = "y")
points(y, pch = 19, col = "gray") 
abline(h = ks_mh$prior_coef) 
par(op)

```

### Model coefficient - Mental Health

```{r}
coef2 <- as.data.frame(t(rbind(ks_mh$prior_coef, ks_mh$prior_coef_se)))
coef2 <- cbind(row.names(coef2),coef2) 
names(coef2) <- c("Predictors", "Coefficient", "Sd.")
flextable(coef2) %>% theme_booktabs()%>% autofit()
```

### Significant Model coefficient - Mental Health

```{r}
coef2$Q97.5 <- coef2$Coefficient + 2 * coef2$Sd.
coef2$Q2.5 <- coef2$Coefficient - 2 * coef2$Sd.
coef2 <- coef2[(coef2$Q97.5 > 0 & coef2$Q2.5 >0)| (coef2$Q97.5 < 0 & coef2$Q2.5 < 0),]
names(coef2) <- c("Predictor", "Coefficient", "Sd.", "Q2.5", "Q97.5")
flextable(coef2) %>% theme_booktabs() %>% autofit()
```


## Refit model for significant predictors - Mental Health

```{r}
smoothing_formula3 <- smoothb ~ MnrPrc_s + MnrPrc_r + MdGrTx_s + MdGrTx_r + Mdn_Inc_s + Mdn_Inc_r  + mn_tmp_s + mn_tmp_r + Tre_Cnp
v <- variogram(smoothing_formula3, newhealth)
```

```{r}
v_fit <- fit.variogram(v, vgm("Wav"))
v_f <- spherical_variogram(v_fit$psill[1], v_fit$psill[2], v_fit$range[2])
h <- seq(0, 200, length = 1000) 
plot(v$dist, v$gamma,  pch = 19, col = "gray",
     xlab = "distance", ylab = "semivariogram", main = "Mental Health")
lines(h, v_f(h))
abline(v = v_fit$range[2], col = "gray")


# check smoothing:
ks_mh_refit <- kriging_smooth_spherical(smoothing_formula3, newhealth)
y <- newhealth$smoothb
op <- par(mfrow = c(1, 2))
plot(ks_mh_refit$smooth, y, ylim = c(2.2,3)); abline(0, 1, col = "red") 
plot(ks_mh_refit$smooth, type = "l", ylab = "y")
points(y, pch = 19, col = "gray") 
abline(h = ks_mh_refit$prior_coef) 
par(op)
```

### Model coefficient - Mental Health

```{r}
coef3 <- as.data.frame(t(rbind(ks_mh_refit$prior_coef, ks_mh_refit$prior_coef_se)))
coef3 <- cbind(row.names(coef3),coef3) 
```


```{r}
coef3$Q97.5 <- coef3$V1 + 2 * coef3$V2
coef3$Q2.5 <- coef3$V1 - 2 * coef3$V2
names(coef3) <- c("Predictor", "Coefficient", "Sd.", "Q2.5", "Q97.5")
# coef3 <- coef3[(coef3$Q97.5 > 0 & coef3$Q2.5 >0)| (coef3$Q97.5 < 0 & coef3$Q2.5 < 0),]
flextable(coef3) %>% theme_booktabs() %>% autofit()
```

# After-Smoothing Map & Kriging Predicted Map for Mental Health
```{r}
newhealth$ks_mh_refit <- ks_mh_refit$smooth
tract7$ks_mh_refit[tract7$GEOID20 %in% newhealth$GEOID20] <- newhealth$ks_mh_refit[newhealth$GEOID20 %in% tract7$GEOID20]
ggplot() + geom_sf(color = 'grey', data = tract7) + geom_sf(aes(fill = ks_mh_refit), data=na.omit(tract7)) + 
  scale_fill_distiller(palette="Blues", trans = "reverse") + labs(title = 'Original Model for Mental Health')
grid <- st_centroid(tract7[is.na(tract7$ks_mh_refit),]) 
pred_MH <-  krige(ks_mh_refit ~1, newhealth, grid$geometry, model = v_fit)
tract7$ks_mh_refit[is.na(tract7$ks_mh_refit)] <- pred_MH$var1.pred
ggplot(data = tract7) + geom_sf(color = 'grey') + geom_sf(aes(fill = ks_mh_refit)) + 
  scale_fill_distiller(palette="Blues", trans = "reverse") + labs(title = 'Refitted Model for Mental Health ')
```


